Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/abstract.tex
large pass for consistency/small bugfixes

In this paper, we consider staged programs written in a typed lambda
calculus with a (``circle'') modality denoting computation in a later
stages and present techniques for splitting them into multi-pass
programs. Such multi-pass programs evaluate in a sequence of
evaluations of un-staged, conventional programs each of which
corresponds to a pass.  Conceived in the 80's but J{\o}rring and
Scherlis, this {\em pass-separation} technique has been studied but
existing techniques are either manual or apply only to simple programs
that lack important features such as recursion and first-class
functions.

@@

Since we build on a lambda calculus, our techniques enable the
programmer to write staged programs by using powerful programming
abstractions in a composable and modular fashion. Our splitting
transformation maps such programs into multi-pass programs that can
run significantly more efficiently by assigning frequent computations
to earlier passes so that they can be re-used (cheaply) later.  Our
experiments based on a prototype implementation show that our
splitting algorithm is able perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

>>>>

In this paper, we consider staged programs written in a typed lambda
calculus with a $\fut$ modality denoting computation in a later
stages and present techniques for splitting them into multi-pass
programs. Such multi-pass programs evaluate in a sequence of
evaluations of unstaged, conventional programs each of which
corresponds to a pass.  Conceived in the 80's by J{\o}rring and
Scherlis, this {\em pass-separation} technique has been studied, but
existing techniques are either manual or applicable only to simple programs
that lack important features such as recursion and first-class
functions.

@@

Since we build on a lambda calculus, our techniques enable the
programmer to write staged programs by using powerful programming
abstractions in a composable and modular fashion. Our splitting
transformation maps such programs into multi-pass programs that can
run significantly more efficiently by assigning frequent computations
to earlier passes so that they can be reused (cheaply) later.  Our
experiments based on a prototype implementation show that our
splitting algorithm is able perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

### 2015-02-28 paper/examples.tex
large pass for consistency/small bugfixes

1`//dot : vec * vec -> $int
fun dot (
  (3`grnd{`1`xliteralone`3`}`1`,`3`grnd{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
  (3`grnd{`1`xliteraltwo`3`}`1`,`3`grnd{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
next{
  2`prev{`1`hold `3`grnd{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

@@

Our next example, \texttt{exp}, is a mainstay of the partial evaluation
literature (for example, in \cite{jones96}). \texttt{exp} recursively computes
$b^e$ using exponentiation by squaring, where $e$ is known at the first stage,
and $b$ is known at the second stage. 
%
\begin{lstlisting} 
1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`grnd{`1`e`3`}`1` : ^int) = 
  if `3`grnd{`1`e == 0`3`}`1` then 
    next{`2`literalone`1`}
  else if `3`grnd {`1`(e mod 2) == 0`3`}`1` then
    exp(next{`2`b*b`1`},`3`grnd{`1`e/literaltwo`3`}`1`)
  else 
    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`grnd{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
\end{lstlisting}
%
Because \texttt{exp} is a recursive function whose conditionals test the parity
of the exponent argument, the sequence of branches taken corresponds exactly to
the binary representation of $e$. Partially evaluating \texttt{exp} with $e$
eliminates all of the conditionals, selecting and expanding the appropriate
branch in each case.

@@

Note that the recursive \texttt{partition} function is defined within a \texttt{grnd}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{grnd} annotations, however, this can be prevented.

>>>>

1`//dot : vec * vec -> $int
fun dot (
  (3`gr{`1`xliteralone`3`}`1`,`3`gr{`1`yliteralone`3`}`1`,next{`2`zliteralone`1`}): vec,
  (3`gr{`1`xliteraltwo`3`}`1`,`3`gr{`1`yliteraltwo`3`}`1`,next{`2`zliteraltwo`1`}): vec) = 
next{
  2`prev{`1`hold `3`gr{`1`(x1*x2) + (y1*y2)`3`}`2`} + (z1*z2)`
1`}`
\end{lstlisting}
%
The body of \texttt{dot} is a \rmint{} term at world \bbtwo\ containing a
\rmint\ computation of \texttt{(x1*x2) + (y1*y2)} which is promoted from world
\bbonep\ to world \bbtwo. We would expect the first stage of the split program
to take the first two coordinates of each vector and perform that first-stage
computation; and the second stage to take the final coordinates and the result
of the first stage, then multiply and add.

@@

Our next example, \texttt{exp}, is a mainstay of the partial evaluation
literature (for example, in \cite{jones96}). \texttt{exp} recursively computes
$b^e$ using exponentiation by squaring, where $e$ is known at the first stage,
and $b$ is known at the second stage. 
%
\begin{lstlisting} 
1`fun exp (next{`2`b`1`} : $`2`int`1` , `3`gr{`1`e`3`}`1` : ^int) = 
  if `3`gr{`1`e == 0`3`}`1` then 
    next{`2`literalone`1`}
  else if `3`gr{`1`(e mod 2) == 0`3`}`1` then
    exp(next{`2`b*b`1`},`3`gr{`1`e/literaltwo`3`}`1`)
  else 
    next{`2`b * prev{`1`exp(next{`2`b*b`1`},`3`gr{`1`(e-1)/literaltwo`3`}`1`)`2`}`1`}`
\end{lstlisting}
%
Because \texttt{exp} is a recursive function whose conditionals test the parity
of the exponent argument, the sequence of branches taken corresponds exactly to
the binary representation of $e$. Partially evaluating \texttt{exp} with $e$
eliminates all of the conditionals, selecting and expanding the appropriate
branch in each case.

@@

Note that the recursive \texttt{partition} function is defined within a \texttt{gr}
annotation.  As explained in \ref{sec:needGround}, defining \texttt{partition}
at \bbonem\ would cause it to split in a way that incurs extra cost at the second stage.
In this case, that cost would be $\Theta(n)$ in the size of the input list,
enough to overpower the asymptotic speedup gained elsewhere.
With \texttt{gr} annotations, however, this can be prevented.

### 2015-02-28 paper/implementation.tex
large pass for consistency/small bugfixes

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngrnd{`1`
  datatype list = Empty | Cons of int * list
  fun partition (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`grnd{`Empty3`}` = 3`grnd{`roll (inj ...)3`}`
val 3`grnd{`Cons3`}` = 3`grnd{`fn (...) => roll (inj ...)3`}`
val 3`grnd{`partition3`}` = 3`grnd{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

>>>>

We allow staging annotations around declarations. For example,
\begin{lstlisting}
3`atsigngr{`1`
  datatype list = Empty | Cons of int * list
  fun partition (...) = ...
`3`}`1`
fun qss (...) = ...`
\end{lstlisting}
declares the \texttt{list} datatype and \texttt{partition} function at world \bbonep,
by elaborating into:
\begin{lstlisting}
1`val 3`gr{`Empty3`}` = 3`gr{`roll (inj ...)3`}`
val 3`gr{`Cons3`}` = 3`gr{`fn (...) => roll (inj ...)3`}`
val 3`gr{`partition3`}` = 3`gr{`fn (...) => ...3`}`
val qss = fn (...) => ...`
\end{lstlisting}

### 2015-02-28 paper/intro.tex
large pass for consistency/small bugfixes

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_{m-1}$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f(x, y_0);  f(x, y_1); \ldots; f(x, y_{m-1})$@
\end{lstlisting}
%
In this common scenario, a multivariate function is applied many times with one
argument fixed, so it is advantageous to find the computations in $f$ that
depend on $x$ and perform them only once. This is the intent of compiler
optimization techniques like loop hoisting and common subexpression elimination,
which identify the computations common to all invocations of $f$, and perform
them once at the beginning of the computation.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_0, y_1, \ldots, y_{m-1})$@ = 
  @$f_v(y_0);  f_v(y_1); \ldots; f_v(y_{m-1})$@.
\end{lstlisting}
eliminating the need to recompute the parts of $f(v,-)$ which do not depend on
the second argument.

A closely related technique, {\em metaprogramming}, allows writing
programs that generate specialized code
(e.g.,~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
enables fine-grained control over specialization by requiring staging
annotations that make explicit the stage of each computation. (In
partial evaluation, we typically assume that a \emph{binding-time
  analysis} has automatically determined which parts of $f$ can be
specialized.)

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and then the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_0, y_1, \ldots, y_{m-1})$@ = 
  let @$fx = f_1(x)$@
  in @$(f_2(fx, y_0);  f_2(fx, y_1)  \ldots f_2(fx, y_{m-1}))$@.
\end{lstlisting}
%
The key difference between partial evaluation (or metaprogramming) and
pass separation is that the latter can be performed without access to
the first argument $x$: $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Note that unlike
precomputation, frequency reduction does not require knowledge of
arguments; if a particular $v$ is known, we can use partial evaluation
to specialize $F_\textrm{multipass}$ to that $v$.

>>>>

Consider a function $F$ which takes an $x$ and $y_1,\dots,y_m$, and computes
$f(x,y_i)$ for each $y_i$:
\begin{lstlisting}
fun @$F(x, y_1, y_2 \ldots, y_m)$@ = 
  @$f(x, y_1);  f(x, y_2); \ldots; f(x, y_m)$@
\end{lstlisting}
%
In this common scenario, a multivariate function is applied many times with one
argument fixed, so it is advantageous to find the computations in $f$ that
depend on $x$ and perform them only once. This is the intent of compiler
optimization techniques like loop hoisting and common subexpression elimination,
which identify the computations common to all invocations of $f$, and perform
them once at the beginning of the computation.

@@

There has been much work on language-based techniques for performing
precomputation. Partial evaluation~\cite{futamura71,jones96} achieves
precomputation by specializing functions to specific arguments. Going back to
our example, if we specialize $f$ to a particular $v$, written $f_v$, such that
$f(v,y) = f_v(y)$, then we can specialize $F$ to $v$ as
\begin{lstlisting}
fun @$F_v(y_1, y_2, \ldots, y_m)$@ = 
  @$f_v(y_1);  f_v(y_2); \ldots; f_v(y_m)$@.
\end{lstlisting}
eliminating the need to compute $m$ times those parts of $f(v,-)$ which do not
depend on the second argument.

A closely related technique, {\em metaprogramming}, allows writing
programs that generate specialized
code~\cite{davies96,Taha97,DP01-modal,NP05-nn}. Metaprogramming
enables fine-grained control over specialization by requiring \emph{staging
annotations} that explicitly mark the stage of each computation. (In
partial evaluation, we typically assume that a \emph{binding-time
  analysis} has automatically determined which parts of $f$ can be
specialized.)

While partial evaluation and metaprogramming have been studied extensively,
frequency reduction techniques have not seen the same attention. In
\cite{JS86-staging}, J{\o}rring and Scherlis proposed {\em pass separation} as a
manual method for {\em splitting} a program into multiple passes.
%
In our example, pass separation would split the function $f$ into two passes
$f_1$ and $f_2$ such that $f(x,y_i) = f_2(f_1(x),y_i)$. Then we evaluate $F$ by
evaluating the first pass on $x$, and then the second pass on each $y_i$.
%
\begin{lstlisting}
fun @$F_\textrm{multipass}(x, y_1, y_2, \ldots, y_m)$@ = 
  let @$z = f_1(x)$@
  in @$f_2(z, y_1);  f_2(z, y_2); \ldots; f_2(z, y_m)$@
\end{lstlisting}
%
The key difference between pass separation and partial evaluation (or
metaprogramming) is that the former can be performed without access to
the first argument $x$; $F_\textrm{multipass}$ works for any $x$,
while $F_v$ is defined only for $x=v$.  Note that unlike
precomputation, frequency reduction does not require knowledge of
arguments; however, if a particular $v$ is known, we can use partial evaluation
to specialize $F_\textrm{multipass}$ to that $v$.

### 2015-02-28 paper/overview.tex
delete obsolete quickselect figure
continuing a pass through the figures
eliminate inconsistent uses of 'part'ition
large pass for consistency/small bugfixes

\input{figures/quickselect-aligned}

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which give a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected linear time.
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate: we wish to be able to map \texttt{qs} to
over another list and do so efficiently.

@@

We can write a staged version of \texttt{qs}, called \texttt{qss},
whose code is shown in \ref{fig:qs-staged}, where the first-stage code
are typeset in red, and the second-stage code are typeset in blue.  To
obtain \texttt{qss} from \texttt{qs}, we first make the input list a
stage-1 and the rank a stage-2 argument.
%
More precisely the input list has type $\curr\mathrm{list}$ (an
integer list ``now''), the rank has type $\fut\rmint$ (an integer in
the ``future''), and the return has type $\fut\rmint$.
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between a stage-1 and stage-2 code, and~$\texttt{g}$, which
marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.
%

Our type system ensures that the staging annotations are consistent,
in the sense that computations marked as first-stage cannot depend on
ones marked as second-stage.
%
The process of adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). We assume that
these annotations have been provided; we do not consider the problem
of generating them. In this example, there are other ways to annotate
\texttt{qs}, but we chose annotations that maximize the work performed
in the first stage

% The $\curr$ and \texttt{grnd} (``ground'') annotations
% further distinguish those first-stage computations which do not
% contain any embedded second-stage code. (Code \emph{inside}
% \texttt{grnd}, like the \texttt{part}ition function, is guaranteed not
% to contain second-stage code; first stage code outside a
% \texttt{grnd}, like \texttt{qsStaged}, may.) We will discuss the
% importance of this additional distinction in \ref{sec:splitting}.

% Ignoring for the moment all \texttt{grnd} annotations, the first-stage code in
% \texttt{qsStaged} looks like \texttt{qSelect}: in the first stage, terms of
% non-circle type are available for immediate use. Indeed, we case on the list as
% usual, and in the \texttt{Cons} branch, \texttt{part}ition it.

@@

\input{figures/quickselect-split-aligned}

>>>>

\input{figures/quickselect}

Suppose that we wish to compute a series of order statistics queries
on a list \texttt{l}. To this end, we can use
quickselect~\cite{Hoare:1961}, which given a list \texttt{l} and a rank
\texttt{k}, returns the element of \texttt{l} with rank \texttt{k}.
As implemented in an ML-like language in \ref{fig:qs-unstaged},
function \texttt{qs} partitions the list by using the first element as
a pivot and then recurs on one of the two resulting sides to find the
desired element based on the relationship of \texttt{k} to the size of
the first half \texttt{i}.  Assuming that the input is uniformly
randomly ordered (which can be achieved by pre-permutation of the
input), function \texttt{qs} runs in expected linear time.
%
Using function \texttt{qs}, we can perform many order statistics
queries, for example with $m$ different ranks
$\mathtt{k_1},\dots,\mathtt{k_m}$ as follows
%
\begin{lstlisting}
qs l @$\mathtt{k_1}$@; qs l @$\mathtt{k_2}$@; @$\ldots$@; qs l @$\mathtt{k_m}$@.
\end{lstlisting}
%
Unfortunately, this approach requires $O(n \cdot m)$ time, making it
expensive to perform order statistics on more than a small number of
times. This is unfortunate: we wish to be able to map \texttt{qs} to
over another list and do so efficiently.

@@

We can write a staged version of \texttt{qs}, called \texttt{qss},
whose code is shown in \ref{fig:qs-staged}, where the first-stage code
are typeset in red, and the second-stage code are typeset in blue.  To
obtain \texttt{qss} from \texttt{qs}, we first make the input list a
stage-1 and the rank a stage-2 argument.
%
More precisely the input list has type $\curr\mathrm{list}$ (an
integer list ``now''), the rank has type $\fut\rmint$ (an integer in
the ``future''), and the return has type $\fut\rmint$.
%
We can then obtain \texttt{qss} by annotating the body of \texttt{qs}
with the staging annotations \texttt{prev} and \texttt{next}, which
transition between a stage-1 and stage-2 code, and~$\texttt{g}$, which
marks ground terms.  
%
We also use a third function
\texttt{hold}, which can be implemented with \texttt{prev} and
\texttt{next}, for promoting a stage-1 integer to a stage-2 integer;
the signature of \texttt{hold} is
% 
\lstinline{1`hold : ^int -> $`2`int`}.

Our type system ensures that the staging annotations are consistent,
in the sense that computations marked as first-stage cannot depend on
ones marked as second-stage.
%
The process of adding staging annotations to unstaged code has been
the subject of extensive research (\ref{sec:related}). We assume that
these annotations have been provided; we do not consider the problem
of generating them. In this example, there are other ways to annotate
\texttt{qs}, but we chose annotations that maximize the work performed
in the first stage.

@@

\input{figures/quickselect-split}

### 2015-02-28 paper/splitting.tex
large pass for consistency/small bugfixes

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` grnd{1+2}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first and second stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

%Masking is not a transformation performed as part of the splitting
%algorithm.  Instead, masking defines an intuitive notion of the first
%stage and second stage ``components'' of residual tables and partial
%values, and it is used to define the requirements of what output
%splitting of \bbonem\ terms must produce.
To provide intuition about the intent of masking, consider the
following residual table and partial value:
\begin{lstlisting}
2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (grnd{7}, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
masking first redacts all second stage (blue) parts, along with the
surrounding \texttt{next} annotations. (This redaction removes the
entire residual table since it only represents second stage
computations.) The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (grnd{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), injL (grnd{7},()))`
\end{lstlisting}
Finally, masking drops \texttt{grnd} annotations yielding:
\begin{lstlisting}
1`((), injL (7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{grnd} blocks
(replacing them with unit), \texttt{next} annotations, and injection tags:
%\begin{lstlisting}
%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
%\end{lstlisting}
%Where the \texttt{grnd} blocks left a hole, we put in a unit values:
\begin{lstlisting}
2`[yhat |-> 1+2] (yhat, ((),yhat))`
\end{lstlisting}
Then the residual table is reified into \texttt{let} bindings,
yielding:
\begin{lstlisting}
2`let yhat = 1+2 in (yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the masking relation is given in
\ref{fig:valMask}.

@@

% Masking assigns ground values to the immediate value
% and likewise assigns references into the residual table to the resumer.
% In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
% Note that the \texttt{grnd} and \texttt{next} annotations are erased.

>>>>

\textbf{\bbonem\ term splitting.}  For any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$) , splitting $e$ (via the new
relation~$\splitonesym$) again yields the pair of monostage terms
$\pipeM c l r$. Evaluating $c$ in the first pass produces the pair
$(i,b)$, where the first element $i$ matches the first stage results
contained in $v$.  For example, in the simple case where $\rtab \xi v$
contains entirely first stage computation, such as:
\begin{lstlisting}
2`[]`1` 3`gr{`1`literalone+literaltwo`3`}`
\end{lstlisting}
the first stage result is the value 3.  The second element of the pair
is the boundary value used as input to the resumer in second pass
evaluation as $\reduce {[b/l]r} {v'}$, where $v'$ is the same as
result produced by second stage computations in $e$.  For example, in
the case where $\rtab \xi v$ contains entirely second stage
computation, such as:
\begin{lstlisting}
2`[yhat |-> 2+3]`1` next{`2`yhat`1`}`
\end{lstlisting}
the second stage result is the value 5. Since in general residual
tables and partial values describe a mixture of both first and second
stage computations, in the subsequent subsection we define an
operation, called {\em masking}, that separates residual tables and
partial values into their first and second stage results. We then
describe the algorithm for splitting both \bbonem\ terms
($\splitonesym$) and \bbtwo\ terms ($\splittwosym$).

@@

%Masking is not a transformation performed as part of the splitting
%algorithm.  Instead, masking defines an intuitive notion of the first
%stage and second stage ``components'' of residual tables and partial
%values, and it is used to define the requirements of what output
%splitting of \bbonem\ terms must produce.
To provide intuition about the intent of masking, consider the
following residual table and partial value:
\begin{lstlisting}
2`[yhat |-> 1+2]`1` (next{`2`yhat`1`}, injL (3`gr{`1`7`3`}`, next{`2`yhat`1`}))`
\end{lstlisting}
To construct the value $i$ representing its first-stage components,
masking first redacts all second stage (blue) parts, along with the
surrounding \texttt{next} annotations. (This redaction removes the
entire residual table since it only represents second stage
computations.) The resulting ``holes'' in the term are replaced with
unit values.
%\begin{lstlisting}
%4`##########`1` (`4`######`1`, injL (gr{7}, `4`######`1`))`
%\end{lstlisting}
%The holes inside the term needs to be filled with something to convey the lack of information, so we use a unit values:
\begin{lstlisting}
1`((), injL (3`gr{`1`7`3`}`,()))`
\end{lstlisting}
Finally, masking drops \texttt{gr} annotations yielding:
\begin{lstlisting}
1`((), injL (7,()))`
\end{lstlisting}
To construct the residual $q$ (corresponding to second stage computations) masking first redacts all \texttt{gr} blocks
(replacing them with unit), \texttt{next} annotations, and injection tags:
%\begin{lstlisting}
%2`[yhat |-> 1+2]`1` (`4`####`2`yhat`4`#`1`, `4`####`1` (`4`####`1`,`4`####`2`yhat`4`#`1`))`
%\end{lstlisting}
%Where the \texttt{gr} blocks left a hole, we put in a unit values:
\begin{lstlisting}
2`[yhat |-> 1+2] (yhat, ((),yhat))`
\end{lstlisting}
Then the residual table is reified into \texttt{let} bindings,
yielding:
\begin{lstlisting}
2`let yhat = 1+2 in (yhat, ((),yhat))`
\end{lstlisting}
A precise definition of the masking relation is given in
\ref{fig:valMask}.

@@

% Masking assigns ground values to the immediate value
% and likewise assigns references into the residual table to the resumer.
% In both cases, the alternate component is assigned to $\tup{}$, to represent trivial information.
% Note that the \texttt{gr} and \texttt{next} annotations are erased.
